prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-07 inst/MyArticle/paper.Rmd
more references. Changes from Paolo.

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all plattforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

Proteins carry out most crucial functions and give structure to living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. Bottom-up mass spectrometric methods, where proteins are specifically and reproducible digested into protein fragments - peptides, are employed to identify and quantify proteins in complex samples containing hundreds to thousand of proteins, [@Bubis2017; @da2020philosopher]. The peptides are first separated by their chemical and physical properties using High-Pressure Liquid Chromatography (HPLC). Afterward, they are ionised, weighted, identified, and quantified using mass spectrometric techniques, e.g., Electro-Spray-Ionization Mass Spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences [@eng2015deeper; @yu2016pipi; @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast; @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath; @demichev2020dia] are counted and aggregated to obtain peptide abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. @rubin1976inference classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

When analyzing parallel-group designs using a single explanatory variable, and contrasting groups, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as `r BiocStyle::Biocpkg("limma")` and `r BiocStyle::Biocpkg("proDA")` allow us to fit a comprehensive variety of models and test various hypotheses; however, in-depth knowledge of design matrix specification using the R formula interface is required [@law2020guide].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, which makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of proteomics data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely (and wildly). However, these packages have in common: they fit linear models either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, where we can add features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modeling of missing observations.

Furthermore, the functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish, robust linear models [@goeminne2020msqrob]. Furthermore, peptide or protein abundances can be scaled and transformed, using robust scaling, transformed using _quantile_ normalization or `r BiocStyle::Biocpkg("vsa")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with other R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We also implemented functionality to streamline the benchmarking of the implemented methods. We use the Ionstar [@shen2018ionstar] dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion (FDP).

@@

We store all the data needed for analysis in a tidy table, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an __R6__[@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, _DiaNN_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats, enables the use of the methods without additional arguments, and preserves the processing tools' differences in the figures and analysis. We show an example code for creating an MSFragger configuration in the Supplement. For popular software like _MaxQuant_, or_MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Furthermore, relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

(ref:LFQData) Class Diagram of classes representing the proteomics data. The LFQData class encapsulates the quantitative proteomics data stored in a tidy table. An instance of the AnalysisTableConfiguration class specifies a mapping of columns in the tidy table. The LFQDataPlotter class and other classes decorate the LFQData class with additional functionality. Similarly, the `LFQDataStats` and`LFQDataSummary`, reference the LFQData class and group methods for variance and sample size estimation or summarizing peptide and protein counts.

@@

In addition, we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r `BiocStyle::Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies using.

(ref:ContrastUML)  UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_, can fit various types of models. The model builder method fit's the statistical model given the data and a strategy. The obtained model can then be used for the analysis of variance or to estimate contrasts. All classes estimating contrasts implement the _Contrasts_ interface. Results of external tools, e.g., SAINTExpress, can be adapted by implementing the Contrasts Interface.

@@

## Dataset for Benchmarking

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, and hence 20 raw files in total. We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata]. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

The peptide abundances (from the _MaxQuant_ _peptide.txt_ file), were $log_2$ transformed and subsequently scaled, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish [@tukey1977exploratory]. Finally, we fitted the fixed effect linear models, the dropout model `r BiocStyle::Biocpkg("proDA")` to protein abundances, the mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model to peptide intensities.

@@

The IonStar [@shen2018ionstar] dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins, the difference $\beta$ between two dilutions should be $\beta = 0$. While for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$.

@@

knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, P - all positive cases (all E. coli proteins), N - all negative cases (all H. sapiens proteins), m- all proteins.")
```

@@

@valikangas2016systematic discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we will be using a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median, and instead of the standard deviation $\tilde{S}$ the median absolute deviation (mad):

@@

Given a linear model $y = \beta X$, we can compute the difference between two groups $\beta_{c}$ by a linear combination $c$ of linear model parameters $\beta$. $c$ is a column vector with as many elements as there are coefficients in the linear model. If $c$ has $0$ for one or more of its rows, then the corresponding coefficient in $\beta$ is not involved in determining the contrast [@ph525xseries].

@@

with $X$ being the design matrix, $\beta$ the model parameters and $c$ an array of weights, defining the contrast (difference among conditions). The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are available in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

To summarize peptide level models to protein models, we applied the method suggested by [@Suomi2017bEnhanced] to use the median scaled $p$-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset containing $163$ proteins and $1258$ peptides. Peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:prepro) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish [@tukey1977exploratory]. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

(ref:scaling) Panel A - Peptide intensity distributions for $20$ samples. For each sample a line with a different colour is shown. Panel B - Peptide intensities for protein _5NTD_ are shown using a line of different colour, and the protein intensity estimate is shown using a fat black line, Panel C - distribution of standard deviations of all proteins in each dilution group (a, b, c, d, e) and overall (all), Panel D - Distribution of protein intensities for protein _5NTD_.

The following code example illustrates how we compute contrast by comparing the protein intensities among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `addContrastResults` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:exampleContrasts). Panel A shows the distribution of the $p$-values, while figure Panel B shows the volcano plot for each comparison.

>>>>

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

Proteins carry out most crucial functions and give structure to living cells. Hence, measuring changes in protein abundance is the subject of active research [@vidova2017review]. Bottom-up mass spectrometric methods, where proteins are specifically and reproducibly digested into protein fragments - peptides, are employed to identify and quantify proteins in complex samples containing hundreds to thousand of proteins [@Bubis2017; @da2020philosopher]. The peptides are first separated by their chemical and physical properties using liquid chromatography (LC). Afterward, they are ionised, weighed, identified, and quantified using mass spectrometric techniques, e.g., electro-spray-ionization mass spectrometry (ESI-MS). Finally, peptide identification is achieved by fragmenting and matching the measured fragment masses to theoretical masses computed from known peptide sequences [@eng2015deeper; @yu2016pipi; @kong2017msfragger]. For quantification, intact peptide ions [@yu2020fast; @Cox2008MaxQuant] or products of peptide ion fragmentation [@rost2014openswath; @demichev2020dia] are counted and aggregated to obtain peptide abundances. Finally, the identified and quantified peptides are assigned to proteins based on protein sequence information.

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. @rubin1976inference classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, a MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), a MNAR mechanism [@mcgurk2020use].

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities, and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

When analyzing parallel-group designs using a single explanatory variable, and contrasting groups, we can use all R packages; but, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as `r BiocStyle::Biocpkg("limma")` and `r BiocStyle::Biocpkg("proDA")` allow us to fit a comprehensive variety of models and test various hypotheses; however, in-depth knowledge of design matrix specification using the R formula interface is required [@law2020guide].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of proteomics data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely. At the same time, they have in common that: they fit linear models either to peptide or protein intensities, determine contrasts among conditions, and afterward apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, where we can add features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modeling of missing observations.

Furthermore, the functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Furthermore, peptide or protein abundances can be scaled and transformed, using robust scaling, _quantile_ normalization or `r BiocStyle::Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with other R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We also implemented functionality and use the Ionstar [@shen2018ionstar] dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion (FDP).

@@

We store all the data needed for analysis in a tidy table, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats, enables the use of the methods without additional arguments, and enables to preserve the differences in the analysis and figures. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (Appendix Figure \@ref(fig:tabCompletion).

(ref:LFQData) Class Diagram of classes representing the proteomics data. The LFQData class encapsulates the quantitative proteomics data stored in a tidy table. An instance of the AnalysisTableConfiguration class specifies a mapping of columns in the tidy table. The LFQDataPlotter class and other classes decorate the LFQData class with additional functionality. Similarly, the `LFQDataStats` and `LFQDataSummary`, reference the `LFQData` class and group methods for variance and sample size estimation or summarizing peptide and protein counts.

@@

In addition, we implement features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r BiocStyle::Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies.

(ref:ContrastUML)  The UML (unified modeling language) diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ (Table \@ref(tab:prolfquaModels)), can fit various types of models. The model builder method fits the statistical model given the data and a strategy. The obtained model can then be used for the analysis of variance or to estimate contrasts. All classes estimating contrasts implement the _Contrasts_ interface. Results of external tools, e.g., SAINTexpress, can be adapted by implementing the Contrasts Interface.

@@

## Dataset for benchmarking

To evaluate the performance of differential expression analysis, we use the IonStar benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_, _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, and hence 20 raw files in total. We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata]. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

The peptide abundances (from the MaxQuant _peptide.txt_ file), were $log_2$ transformed and subsequently scaled, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r BiocStyle::Biocpkg("proDA")` to protein abundances, the mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model to peptide intensities.

@@

The IonStar dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins, the difference $\beta$ between two dilutions should be $\beta = 0$. While for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$.

@@

knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, TN - true negatives, P - all positive cases (all $E.~coli$ proteins), N - all negative cases (all $H.~sapiens$ proteins), m - all proteins.")
```

@@

@valikangas2016systematic discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we use a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median, and instead of the standard deviation $\tilde{S}$ the median absolute deviation (mad):

@@

Given a linear model $y = \beta X$, we can compute the difference between two groups $\beta_{c}$ by a linear combination $c$ of linear model parameters $\beta$, where $c$ is a column vector with as many elements as there are coefficients $\beta$ in the linear model. If $c$ has $0$ for one or more of its rows, then the corresponding coefficient in $\beta$ is not involved in determining the contrast [@ph525xseries].

@@

with $X$ being the design matrix, $\beta$ the model parameters and $c$ an array of weights, defining the contrast (difference among conditions). The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")`  and used the Satterthwaite [@Kuznetsova2017lmerTest] method to estimate the denominator degrees of freedom. These methods are available in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

To summarize peptide level models to protein models, we applied the method suggested by @Suomi2017bEnhanced to use the median scaled $p$-value of the peptide models and the cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset containing $163$ proteins and $1258$ peptides. Peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:prepro) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

(ref:scaling) Panel A - Peptide intensity distributions for $20$ samples. For each sample a line with a different colour is shown. Panel B - Peptide intensities for protein _5NTD_ are shown using a line of different colour, and the protein intensity estimate is shown using a fat black line, Panel C - distribution of standard deviations of all proteins in each dilution group ($a$, $b$, $c$, $d$, $e$) and overall (all), Panel D - Distribution of protein intensities for protein _5NTD_.

The following code example illustrates how we compute contrast by comparing the protein intensities among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `addContrastResults` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:exampleContrasts). Panel A shows the distribution of the $p$-values, while Panel B shows the volcano plot for each comparison.
